---
title: "Visualising drivers of variation in federated omic studies"
author: "Elisabetta Sciacca"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Visualising drivers of variation in federated omic studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)

# DSLite demo chunks run only if all required packages are available
run_dslite <- requireNamespace("DSLite",            quietly = TRUE) &&
              requireNamespace("DSI",               quietly = TRUE) &&
              requireNamespace("dsBaseClient",       quietly = TRUE) &&
              requireNamespace("dsSwissKnifeClient", quietly = TRUE) &&
              requireNamespace("dsPCAdrivers",       quietly = TRUE)
```

# Introduction

`ds.plotDrivers()` generates a tile plot of statistical associations between
principal components (PCs) and clinical or technical variables in a federated
[DataSHIELD](https://www.datashield.org) setting. PCA is computed internally
via `dsSwissKnifeClient::dssPrincomp()` — no pre-existing PCA object is
required. Only aggregate statistics (p values) leave the servers, preserving
individual-level data privacy.

The function supports two modes of analysis:

- **Federated (combined) analysis** (`type = "combine"`): a single PCA is
  computed on the pooled covariance matrix across all sites. All sites share the
  same PC loadings, making cross-site p value combination statistically
  meaningful. P values are combined using Fisher's method and a single heatmap
  is returned.
- **Site-specific analysis** (`type = "split"`): a separate PCA is computed
  independently on each server. Because the resulting PCs may differ across
  sites, p values are not combined, and a faceted heatmap (one panel per site)
  is returned.

The server-side package `dsPCAdrivers` must be installed on each DataSHIELD
server. Source code is available at
<https://github.com/elisabettasciacca/dsPCAdrivers>.


# Installation

```{r install, eval = FALSE}
# Install the client package
devtools::install_github("elisabettasciacca/dsPCAdriversClient")

# The server-side package must be installed on each DataSHIELD server
# by the server administrator:
# devtools::install_github("elisabettasciacca/dsPCAdrivers")
```


---


# Part 1: Local Demonstration with DSLite

[DSLite](https://cran.r-project.org/package=DSLite) is an in-memory DataSHIELD
server that runs entirely within a single R session. It implements the same DSI
interface as a real Opal server, making it ideal for development, testing, and
reproducing examples without requiring network access.

This section uses the `site_data` example dataset included in the package,
which simulates two-site federated RNA-seq data with a known biological
structure: disease cases upregulate the first 15 genes, and sequencing batch
introduces a shift in the first 8 genes.

## Setting up the DSLite environment

```{r dslite-setup, eval = run_dslite}
library(DSLite)
library(DSI)
library(dsPCAdrivers)
library(dsPCAdriversClient)
library(dsBaseClient)
library(dsSwissKnifeClient)
library(dsPCAdriversClient)

# Load simulated example data (2 sites: 60 and 50 samples, 30 genes each)
data(site_data)

# Create a DSLite server holding all four tables
dslite.server <- DSLite::newDSLiteServer(
  tables = list(
    site1_expr     = site_data$site1$expr,
    site2_expr     = site_data$site2$expr,
    site1_clinical = site_data$site1$clinical,
    site2_clinical = site_data$site2$clinical
  )
)

# Load dsBase and dsSwissKnife methods, then register the server-side function
dslite.server$config(DSLite::defaultDSConfiguration(include = c("dsBase", "dsSwissKnife")))
dslite.server$aggregateMethod("plotDriversDS", dsPCAdrivers::plotDriversDS)

# Relax privacy filters for local testing only.
# These settings must NEVER be used with real sensitive data.
# nfilter.glm = 1.0 is required here because the gene/sample ratio (30/60 = 0.5)
# exceeds the default threshold of 0.33 used by ds.cov internally.
options(
  datashield.privacyControlLevel = "permissive",
  nfilter.tab     = 3,
  nfilter.subset  = 3,
  nfilter.glm     = 1.0,
  nfilter.string  = 80,
  nfilter.kNN     = 3
)
```

## Connect and assign data

```{r dslite-login, eval = run_dslite}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "site1", url = "dslite.server",
               table = "site1_expr", driver = "DSLiteDriver")
builder$append(server = "site2", url = "dslite.server",
               table = "site2_expr", driver = "DSLiteDriver")

conns <- DSI::datashield.login(logins = builder$build(), assign = FALSE)

# Assign expression matrices (samples x genes — already in the correct orientation)
DSI::datashield.assign.table(
  conns  = conns,
  symbol = "expression_data",
  table  = c(site1 = "site1_expr", site2 = "site2_expr")
)

# Assign clinical metadata
DSI::datashield.assign.table(
  conns  = conns,
  symbol = "clinical_vars",
  table  = c(site1 = "site1_clinical", site2 = "site2_clinical")
)
```

## Federated analysis

A single PCA is computed on the pooled covariance matrix. Each site then
projects its samples onto the shared PC axes. P values from both sites are
combined using Fisher's method, yielding a single heatmap.

Because the pooled covariance matrix is already n-weighted, an additional
sample-size weight at the combination step is deliberately omitted — it would
double-count the contribution of larger sites.

After the function returns, the PC scores are retained on each server as
`"expression_data_scores"` and are available for downstream analyses
(e.g. biplots).

```{r dslite-combine, eval = run_dslite}
p_combine <- ds.plotDrivers(
  data        = "expression_data",
  vars        = "clinical_vars",
  datasources = conns,
  type        = "combine",
  n_pc        = 5,
  sig_cutoff  = 0.05,
  title       = "Federated PCA — combined associations"
)
print(p_combine)
```

## Site-specific analysis

Each site computes its own PCA independently. The resulting PCs may differ in
direction and magnitude across sites, so p values are not combined. The output
is a faceted heatmap with one panel per site.

```{r dslite-split, eval = run_dslite}
p_split <- ds.plotDrivers(
  data        = "expression_data",
  vars        = "clinical_vars",
  datasources = conns,
  type        = "split",
  n_pc        = 5,
  sig_cutoff  = 0.05,
  title       = "Site-specific PCA — per-site associations"
)
print(p_split)
```

## Cleanup

```{r dslite-logout, eval = run_dslite}
DSI::datashield.logout(conns)
```


---


# Part 2: Real Opal Infrastructure

This section illustrates the same workflow on a real DataSHIELD/Opal
deployment. All code chunks are non-evaluated (`eval = FALSE`) and serve as a
template. Plot outputs shown below are representative examples.

## Prerequisites

`ds.plotDrivers()` requires two objects to already be assigned on each server
before the function is called:

- an expression matrix in **samples × features** orientation (note that omic
  matrices are typically stored as features × samples and must be transposed
  beforehand, e.g. via `datashield.assign.expr(conns, "expr_t", quote(t(expr)))`)
- a data frame of variables to associate with the PCs (e.g. clinical metadata)

## Connect to Opal servers

```{r opal-connect, eval = FALSE}
library(DSI)
library(DSOpal)
library(dsPCAdriversClient)

builder <- DSI::newDSLoginBuilder()
builder$append(
  server   = "site1",
  url      = "https://site1.example.org",
  user     = "username",
  password = "password",
  driver   = "OpalDriver"
)
builder$append(
  server   = "site2",
  url      = "https://site2.example.org",
  user     = "username",
  password = "password",
  driver   = "OpalDriver"
)

conns <- DSI::datashield.login(logins = builder$build(), assign = FALSE)
```

## Federated analysis

```{r opal-combine, eval = FALSE}
p <- ds.plotDrivers(
  data        = "expression_data",  # already assigned on each server
  vars        = "clinical_vars",    # already assigned on each server
  datasources = conns,
  type        = "combine",
  n_pc        = 10,
  sig_cutoff  = 0.05
)
print(p)
```

<!-- INSERT PLOT: federated analysis output -->

## Site-specific analysis

```{r opal-split, eval = FALSE}
p <- ds.plotDrivers(
  data        = "expression_data",
  vars        = "clinical_vars",
  datasources = conns,
  type        = "split",
  n_pc        = 10,
  sig_cutoff  = 0.05
)
print(p)
```

<!-- INSERT PLOT: site-specific analysis output -->

## Parametric vs. non-parametric associations

By default, `ds.plotDrivers()` uses parametric tests (Pearson correlation and
ANOVA). This is the methodologically preferred approach for two reasons.

**Linearity.** PCA is fundamentally a linear dimensionality reduction technique.
It defines orthogonal axes based on linear covariance, so evaluating these axes
with linear models perfectly aligns with the mathematical nature of the
components.

**Normality and power.** Because PCs are linear combinations of numerous omic
features, their distributions tend toward normality by the Central Limit
Theorem, satisfying the core assumption of parametric tests. Parametric tests
also provide greater statistical power, which is critical when applying
stringent p value corrections across many clinical features.

You should switch to rank-based non-parametric tests (Spearman correlation and
Kruskal-Wallis) with `parametric = FALSE` in the following scenarios:

- **Ordinal variables**: ranked categories such as tumour stage I–IV or symptom
  severity, where the distance between levels is not strictly uniform.
- **Extreme outliers**: highly skewed continuous variables such as untransformed
  survival times or blood marker levels that would disproportionately influence
  a linear fit.

```{r opal-parametric, eval = FALSE}
# Non-parametric (Spearman / Kruskal-Wallis)
p <- ds.plotDrivers(
  data        = "expression_data",
  vars        = "clinical_vars",
  datasources = conns,
  parametric  = FALSE,
  n_pc        = 10
)
print(p)
```

## Multiple testing correction

No correction is applied by default. With `p_adj`, adjustment is applied after
Fisher's combination when `type = "combine"`, and independently per site when
`type = "split"`.

```{r opal-padj, eval = FALSE}
p <- ds.plotDrivers(
  data        = "expression_data",
  vars        = "clinical_vars",
  datasources = conns,
  p_adj       = "BH",    # Benjamini-Hochberg FDR correction
  sig_cutoff  = 0.05
)
print(p)
# Other options: "bonferroni", "holm", "hochberg", "hommel", "BY", "fdr"
```

## Plot customisation

```{r opal-customise, eval = FALSE}
p <- ds.plotDrivers(
  data                 = "expression_data",
  vars                 = "clinical_vars",
  datasources          = conns,
  n_pc                 = 10,
  sig_cutoff           = 0.01,
  label                = TRUE,         # print -log10(p) values on tiles
  max_col              = 15,           # fix colour scale maximum
  title                = "Drivers of variation",
  legend               = "bottom",
  transpose_plot       = TRUE,         # PCs on y-axis, variables on x-axis
  drop_insignificant_x = TRUE,         # remove PCs with no significant associations
  drop_insignificant_y = TRUE          # remove variables with no significant associations
)
print(p)
```

## Returning results as a data frame

Set `return_pvalues = TRUE` to retrieve the underlying data instead of a plot,
useful for further statistical analysis or custom visualisation.

```{r opal-return, eval = FALSE}
res <- ds.plotDrivers(
  data           = "expression_data",
  vars           = "clinical_vars",
  datasources    = conns,
  return_pvalues = TRUE
)

# Columns: Feature, PC, pvalue, Association (-log10 p), Significant
head(res$results)

# Significant associations only
res$results[res$results$Significant, ]

# Metadata (n per site, combination method, etc.)
str(res$metadata)
```

## Cleanup

```{r opal-logout, eval = FALSE}
DSI::datashield.logout(conns)
```


---


# Interpreting the heatmap

Tile colour encodes association strength as $-\log_{10}(p)$: white indicates no
evidence (p ≈ 1) and dark blue indicates strong evidence (p << 0.05). A black
border marks tiles where p ≤ `sig_cutoff`.

**Technical confounders** tend to produce strong associations with the earliest
PCs (PC1, PC2), which capture the largest sources of variance in the data. If
batch, processing date, or library size dominate the first components, batch
correction should be considered before downstream analysis.

**Biological drivers** typically emerge in later PCs and reflect lower-variance
but biologically meaningful signals such as disease status, age, or treatment
response.


# Privacy guarantees

The following information is returned to the client:

- p values (aggregate statistics)
- number of observations per site
- PC and variable names

The following information never leaves the server:

- individual-level PC scores
- raw expression or omic data
- individual variable values
- covariance matrices


# Session info

```{r session-info}
sessionInfo()
```
