---
title: "Plot variables most influencing your Principal Components"
author: "Elisabetta Sciacca"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Plot variables most influencing your Principal Components}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `ds.plotDrivers` function enables privacy-preserving visualization of associations between principal components (PCs) and clinical/technical variables in federated data analysis studies.

## Key Features

- **Privacy-preserving**: Only aggregate statistics (p-values) leave the server
- **Automatically scales**: Works identically for one site or many sites
- **Weighted meta-analysis**: Larger sites get more weight (default)
- **Multi-package support**: Works with dsML and dsSwissKnife PCA outputs
- **Flexible**: Supports blocking variables, multiple testing correction, and various customizations

# Installation

```{r install, eval=FALSE}
# Install client package
devtools::install_github("your-repo/dsPCAdriversClient")

# Server-side package must be installed on DataSHIELD servers
# Contact your DataSHIELD administrator
```

# Basic Workflow

The function works the same way whether you have data on one site or multiple sites.

## Setup Connection(s)

```{r connect}
library(DSI)
library(DSOpal)
library(dsPCAdriversClient)

# Connect to one or more DataSHIELD servers
builder <- DSI::newDSLoginBuilder()

# Add your site(s)
builder$append(
  server = "site1",
  url = "https://site1.org",
  user = "username",
  password = "password",
  driver = "OpalDriver"
)

# Add more sites if available
# builder$append(server = "site2", url = "https://site2.org", ...)
# builder$append(server = "site3", url = "https://site3.org", ...)

logindata <- builder$build()
conns <- DSI::datashield.login(logins = logindata, assign = TRUE)
```

## Load Data

```{r load_data}
# Assign expression data and clinical variables
# This happens on each connected server
datashield.assign.table(
  conns = conns,
  symbol = "expression_data",
  table = "PROJECT.RNASEQ_COUNTS"
)

datashield.assign.table(
  conns = conns,
  symbol = "clinical_vars",
  table = "PROJECT.CLINICAL_DATA"
)
```

## Compute Federated PCA with dsML

dsML computes **federated PCA** - the principal components are common across all sites, as if all data were in one location.

```{r federated_pca}
library(dsMLClient)

# Compute federated PCA across ALL connected sites
# (works identically whether you have 1 site or 10 sites)
datashield.assign.expr(
  conns = conns,
  symbol = "pca_result",
  expr = quote(ds.PCA(
    x = "expression_data",
    ncomp = 10,
    scale = TRUE,
    center = TRUE
  ))
)
```

## Visualize PC Drivers

```{r plot_drivers}
# Plot associations between PCs and clinical variables
# Automatically combines results across all sites
p <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,     # All sites
  n_pc = 10,
  sig_cutoff = 0.05        # Significance threshold
)

print(p)
```

**What happens:**
- If `conns` has 1 site: plots associations for that site
- If `conns` has 2+ sites: computes associations at each site, combines p-values using weighted Fisher's method, plots combined results

**No p-value adjustment by default** - to apply correction, add `p_adj = "BH"` or another method.

# Advanced Features

## Multiple Testing Correction

By default, **no correction** is applied. Add correction if testing many associations:

```{r p_adjustment}
p <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  n_pc = 10,
  p_adj = "BH",           # Benjamini-Hochberg FDR correction
  sig_cutoff = 0.05
)

# Other options: "bonferroni", "holm", "hochberg", "BY", "fdr"
```

## Weighting in Multi-Site Analysis

When combining results from multiple sites, by default larger sites get more weight:

```{r weighting}
# Default: weighted by sample size
p_weighted <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  meta_method = "weighted_fisher"    # Default
)

# Alternative: equal weight for all sites
p_equal <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  meta_method = "fisher"             # Equal weights
)
```

**Weighted Fisher's method** (default):
- Sites weighted by sample size
- A site with 1000 samples has more influence than a site with 100 samples
- More statistically efficient
- Recommended in most cases

## Blocking Variables for Longitudinal Data

When subjects are measured at multiple timepoints, use blocking:

```{r blocking}
# Clinical data structure:
# - subject_id: repeated for each timepoint
# - timepoint: measurement occasion (1, 2, 3, ...)
# - sex, age: constant within subject
# - treatment, dose: may vary by timepoint

p <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  block = "subject_id",           # Account for within-subject correlation
  n_pc = 10
)
```

**How blocking works:**
- Regresses out the blocking variable before testing associations
- All other variables (sex, age, treatment, etc.) are tested with this adjustment
- Use for repeated measures, family studies, paired samples, etc.

## Customizing the Plot

```{r customize}
p <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  n_pc = 10,
  label = TRUE,                    # Show -log(p) values on tiles
  sig_cutoff = 0.01,               # Stricter significance threshold
  max_col = 15,                    # Fix color scale maximum
  title = "PC Drivers Analysis",
  legend = "bottom",
  transpose_plot = TRUE,           # PCs on y-axis, variables on x-axis
  drop_insignificant_x = TRUE,     # Remove PCs with no significant associations
  drop_insignificant_y = TRUE      # Remove variables with no significant associations
)

print(p)
```

## Parametric vs Non-Parametric Tests

```{r parametric}
# Parametric tests (default)
# - Continuous variables: Pearson correlation
# - Categorical variables: ANOVA F-test
p_param <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  parametric = TRUE
)

# Non-parametric tests
# - Continuous variables: Spearman correlation
# - Categorical variables: Kruskal-Wallis test
p_nonparam <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  parametric = FALSE
)
```

**Use non-parametric when:**
- Data are skewed or have outliers
- Small sample sizes
- Ordinal variables
- Assumptions of normality violated

## Getting Results as Data

```{r return_data}
# Return results dataframe instead of plot
results <- ds.plotDrivers(
  pcs = "pca_result",
  vars = "clinical_vars",
  datasources = conns,
  return_data = TRUE
)

# Examine results
head(results$results)

# Significant associations only
significant <- results$results[results$results$Significant, ]
print(significant)

# Metadata
str(results$metadata)
results$metadata$n_sites          # Number of sites
results$metadata$total_observations
results$metadata$site_metadata    # Per-site information
```

# Alternative: Site-Specific PCA

If you computed PCA **separately** at each site (not federated), you cannot combine results with meta-analysis because PCs have different meanings at each site.

Instead, create comparison plots:

```{r site_specific}
library(dsSwissKnifeClient)

# Compute PCA at each site independently
# (dsSwissKnife does NOT do federated PCA)
datashield.assign.expr(
  conns = conns,
  symbol = "site_pca",
  expr = quote(dssPrincomp(
    x = "expression_data",
    scale = TRUE,
    center = TRUE
  ))
)

# Create side-by-side comparison
p_comparison <- ds.plotDrivers(
  pcs = "site_pca",
  vars = "clinical_vars",
  datasources = conns,
  federated_pca = FALSE,    # PCs are site-specific
  n_pc = 5,
  title = "Site-by-Site Comparison"
)

print(p_comparison)
```

This creates a **faceted plot** with one panel per site, useful for:
- Comparing patterns across sites
- Identifying site-specific effects
- Quality control

# Interpretation Guide

## Understanding the Heatmap

- **Color intensity**: Strength of statistical evidence (-log₁₀ p-value)
  - White = no evidence (p ≈ 1)
  - Light blue = weak evidence (p between 0.05 and 0.1)
  - Medium blue = moderate evidence (p between 0.01 and 0.05)
  - Dark blue = strong evidence (p < 0.01)

- **Black border**: Association passes significance threshold (p ≤ sig_cutoff)

- **Missing tiles**: Not enough data to test (< 3 complete observations)

## What the Plot Tells You

### Technical Confounders
Strong associations with early PCs (PC1, PC2) often indicate:
- **Batch effects**: Sequencing batch, processing date
- **Technical quality**: RNA quality (RIN), library size
- **Platform effects**: Sequencing center, instrument

**Action:** Consider batch correction if technical variables dominate

### Biological Drivers
Associations with biological variables reveal:
- Which PCs capture biological variation
- Which variables explain most variation
- Unexpected relationships (may indicate data issues or novel biology)

### Common Patterns

**Pattern 1: Batch dominates PC1**
```
      PC1  PC2  PC3
batch ███  ▓▓▓  ░░░
age   ░░░  ▓▓▓  ███
sex   ░░░  ░░░  ▓▓▓
```
→ Batch correction needed

**Pattern 2: Disease status spans multiple PCs**
```
         PC1  PC2  PC3  PC4
disease  ███  ███  ▓▓▓  ▓▓▓
```
→ Disease explains substantial variation, use multiple PCs in downstream analysis

**Pattern 3: No associations**
```
      PC1  PC2  PC3
var1  ░░░  ░░░  ░░░
var2  ░░░  ░░░  ░░░
```
→ May need more PCs, different variables, or larger sample

# Practical Example

Complete analysis workflow:

```{r complete_workflow}
# 1. Connect to sites
conns <- datashield.login(...)

# 2. Load and prepare data
datashield.assign.table(conns, "counts", "PROJECT.COUNTS")
datashield.assign.table(conns, "clinical", "PROJECT.CLINICAL")

# 3. Compute federated PCA
library(dsMLClient)
datashield.assign.expr(conns, "pca",
  quote(ds.PCA("counts", ncomp = 20, scale = TRUE, center = TRUE)))

# 4. Check for technical confounders
p_tech <- ds.plotDrivers(
  pcs = "pca",
  vars = "technical_vars",  # batch, date, RIN, etc.
  datasources = conns,
  n_pc = 10,
  p_adj = "BH",
  title = "Technical Confounders Check"
)

# 5. Analyze biological associations
p_bio <- ds.plotDrivers(
  pcs = "pca",
  vars = "clinical",
  datasources = conns,
  n_pc = 10,
  p_adj = "BH",
  title = "Biological Drivers"
)

# 6. Identify PCs associated with outcome
results <- ds.plotDrivers(
  pcs = "pca",
  vars = "clinical",
  datasources = conns,
  return_data = TRUE
)

disease_pcs <- results$results %>%
  filter(Feature == "disease_status" & Significant) %>%
  pull(PC) %>%
  unique()

print(paste("Disease-associated PCs:", paste(disease_pcs, collapse = ", ")))

# 7. Use these PCs in downstream analysis
# (differential expression, prediction, etc.)
```

# Troubleshooting

## Error: Object not found on server

```{r troubleshoot1, eval=FALSE}
# Check available objects
datashield.symbols(conns)

# Or use aggregate to list
datashield.aggregate(conns, call("ls"))
```

## Error: PCs not federated

This occurs when PC names differ across sites, indicating non-federated PCA.

**Solution:**
```{r troubleshoot2, eval=FALSE}
# Use federated_pca = FALSE for comparison
ds.plotDrivers(..., federated_pca = FALSE)

# Or ensure using dsML for federated PCA
```

## Warning: Variables differ across sites

Some variables may not be available at all sites.

```{r troubleshoot3, eval=FALSE}
# Check which variables are available at each site
for (i in seq_along(conns)) {
  vars <- datashield.aggregate(
    conns[i],
    call("colnames", as.symbol("clinical_vars"))
  )
  cat("Site", i, ":", paste(vars[[1]], collapse = ", "), "\n")
}
```

Meta-analysis includes only common variables.

## No significant associations

**Possible reasons:**
1. Sample size too small → combine more sites
2. Need more PCs → increase `n_pc`
3. Variables not biologically relevant
4. Too strict multiple testing → try `p_adj = "BH"` instead of "bonferroni"

# Technical Details

## Statistical Tests

### Continuous Variables
- **Parametric**: Pearson correlation
- **Non-parametric**: Spearman correlation  
- **With blocking**: Partial correlation (residuals after regressing out block)

### Categorical Variables
- **Parametric**: ANOVA F-test
- **Non-parametric**: Kruskal-Wallis test
- **With blocking**: Repeated measures ANOVA

### Multi-Site Combination

**Weighted Fisher's method** (default):
- Test statistic: Σ(w_i × -2 × log(p_i))
- Weights: w_i = n_i / Σn_i (sample size proportion)
- Effective df: (Σw_i)² / Σ(w_i²) × 2
- More power when sample sizes differ

**Standard Fisher's method**:
- Test statistic: -2 × Σlog(p_i)
- Degrees of freedom: 2k (k = number of sites)
- Equal weight for all sites

## Privacy Guarantees

**What leaves the server:**
- ✓ P-values (aggregate statistics)
- ✓ Number of observations
- ✓ PC and variable names

**What stays on server:**
- ✗ Individual PC scores
- ✗ Individual variable values
- ✗ Raw expression data
- ✗ Covariance matrices

# References

1. Fisher, R.A. (1932). Statistical Methods for Research Workers. Oliver & Boyd.
2. Kost, J.T. & McDermott, M.P. (2002). Combining dependent P-values. Statistics & Probability Letters, 60(2), 183-190.
3. DataSHIELD: <https://www.datashield.org>
4. dsML package: <https://github.com/isglobal-brge/dsML>
5. dsSwissKnife package: <https://github.com/sib-swiss/dsSwissKnife>

# Session Info

```{r session_info}
sessionInfo()
```

# Cleanup

```{r cleanup}
datashield.logout(conns)
```
